---
title: "Analyzing Titanic Tragedy"
output: 
  html_document: default
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(stargazer)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Historical Background

On April 15, 1912, the RMS Titanic sank in the North Atlantic Ocean after striking an iceberg during her maiden voyage from Southampton to New York City. Of the estimated 2,224 passengers and crew aboard, more than 1,500 died, making it one of the deadliest maritime disasters in modern history.

The tragedy led to major improvements in maritime safety regulations and highlighted significant social inequalities of the early 20th century, as passenger survival rates varied dramatically by social class and gender.

## The Data

This dataset includes Survival status, Passenger Class, Sex, Age, Siblings, Parents, Fare.

```{r}
download.file("https://stanford.io/2O9RUCF", destfile = "titanic_temp.csv", mode = "wb")
titanic <- read.csv("titanic_temp.csv")
head(titanic)
```

### Part 1: Data Exploration

```{r}
titanic <- titanic %>% 
  mutate(
    survived = factor(Survived, levels = c(0, 1), labels = c("No", "Yes")),
    Age_group = cut(Age, breaks = c(0, 18, 30, 55, Inf),
                    labels = c("Child (0-18)", "Young Adult (19-30)", "Adult (31-55)", "Elderly (56+)"),
                    right = FALSE)
  )

# Overall survival
ggplot(titanic, aes(x = survived, fill = survived)) + 
  geom_bar() + 
  labs(title = "Overall Survival Count", x = "Survived", y = "Count") + 
  theme_minimal()
```

```{r}
# By class
ggplot(titanic, aes(x = factor(Pclass), fill = survived)) + 
  geom_bar(position = "fill") + 
  labs(title = "Survival by Passenger Class", x = "Class", y = "Proportion") + 
  theme_minimal()
```

```{r}
# By sex
ggplot(titanic, aes(x = Sex, fill = survived)) + 
  geom_bar(position = "fill") + 
  labs(title = "Survival by Sex", x = "Sex", y = "Proportion") + 
  theme_minimal()
```

```{r}
# By age group
ggplot(titanic, aes(x = Age_group, fill = survived)) + 
  geom_bar(position = "fill") + 
  labs(title = "Survival by Age Group", x = "Age Group", y = "Proportion") + 
  theme_minimal()
```

### Part 2: Linear Probability Model

```{r}
model <- lm(Survived ~ Sex + factor(Pclass) + Age, data = titanic)
summary(model)
stargazer(model, type = "text", title = "LPM Results")
```

### Part 3: Logit Model

```{r}
model_1 <- glm(Survived ~ Sex + factor(Pclass) + Age, data = titanic, family = binomial)
summary(model_1)
stargazer(model_1, type = "text", title = "Logit Results")
odds_ratios <- exp(coef(model_1))
data.frame(Variable = names(odds_ratios), Odds_Ratio = as.numeric(odds_ratios))
```

### Part 4: Policy Analysis

The analysis reveals clear patterns in Titanic survival rates driven by social class, gender, and age — reflecting the tragic inequalities of the era.
