---
title: "Analyzing Titanic Tragedy"
output: 
  html_document: default
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(stargazer)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Historical Background

On April 15, 1912, the RMS Titanic sank in the North Atlantic Ocean after striking an iceberg during her maiden voyage from Southampton to New York City. Of the estimated 2,224 passengers and crew aboard, more than 1,500 died, making it one of the deadliest maritime disasters in modern history.

The tragedy led to major improvements in maritime safety regulations and highlighted significant social inequalities of the early 20th century, as passenger survival rates varied dramatically by social class and gender.

## The Data

We're going to look at passenger records from the titanic. This dataset includes includes:

- Survival status (1 = Survived, 0 = Did not survive)
- Passenger Class (1st, 2nd, or 3rd class)
- Sex (Male/Female)
- Age
- Number of Siblings (of the passenger) Aboard
- Number of Parents (of the passenger) Aboard.PHONY: help deps render serve open clean

R_LIB := $(HOME)/R/library
PKGS := rmarkdown knitr dplyr ggplot2 stargazer

help:
    @echo "Targets: deps render serve open clean"

deps:
    sudo apt update
    sudo apt install -y pandoc libcurl4-openssl-dev libssl-dev libxml2-dev build-essential || true
    mkdir -p $(R_LIB)
    Rscript -e '.libPaths(c("$(R_LIB)", .libPaths())); install.packages(c($(PKGS)), repos="https://cran.rstudio.com/", Ncpus = parallel::detectCores())'

render:
    Rscript -e '.libPaths(c("$(R_LIB)", .libPaths())); rmarkdown::render("Titanic_Exercise Solved.Rmd", output_format="html_document")'

serve:
    python3 -m http.server 8000 --directory /workspaces/Analyzing-Titanic-Tragedy &

open:
    "$$BROWSER" "http://localhost:8000/Titanic_Exercise%20Solved.html"

clean:
    rm -f *.html titanic_temp.csv
- Fare paid

Start by downloading the dataset and getting acquainted with it.
```{r}
# Download data
download.file("https://stanford.io/2O9RUCF", 
              destfile = "titanic_temp.csv", 
              mode = "wb")

titanic <- read.csv("titanic_temp.csv")
```


### Part 1: Data Exploration

1. Create appropriate visualizations showing:
   - Overall survival rate
   - Survival rates by passenger class
   - Survival rates by sex
   - Survival rates by age groups
2. What factors seem to have the strongest relationship with survival?
3. Were these patterns what you expected given the historical context?

```{r}
# Convert survived to factor with labels
titanic <- titanic %>% 
  mutate(
    survived = factor(Survived, levels= c(0,1), labels= c("No","Yes")),
    
  )

# Define Age Groups (Children., Young Adults, Adults, Elderly)

titanic <- titanic %>% 
  mutate(Age_group = cut(Age, 
                         breaks = c(0,18,30,55,Inf),
                         labels= c("Child (0-18)", "Young Adult (19-30)", "Adult (31-55)", "Elderly (56+)"),
                         right = FALSE))

# Plotting
# 1. Overall survival rate
ggplot(titanic, aes(x = survived, fill =  survived))+
  geom_bar()+
  labs(title = "Overall Survival Count", x= "Survived", y =  "Count")+
  theme_minimal()
  


```

```{r}
#2. Survival rates by passenger class
ggplot(titanic, aes(x = Pclass, fill = survived)) +
  geom_bar(position = "fill") + # 'position="fill"' shows proportions
  labs(title = "Survival Rate by Passenger Class", x = "Passenger Class", y = "Proportion", fill = "Survived") +
  theme_minimal()
```


```{r}
# 3. Survival rates by sex
ggplot(titanic, aes(x = Sex, fill = survived))+
  geom_bar(position = "fill") + # 'position="fill"' shows proportions
  labs(title = "Survival Rate by Sex", x = "Sex", y = "Proportion", fill = "Survived") +
  theme_minimal()
  
```


```{r}
# 4. Survival rates by age groups
ggplot(titanic, aes(x = Age_group, fill = survived)) +
  geom_bar(position = "fill") + # 'position="fill"' shows proportions
  labs(title = "Survival Rate by Age Group", x = "Age Group", y = "Proportion", fill = "Survived") +
  theme_minimal()
```

What factors seem to have the strongest relationship with survival?
Sex: This is the a dominant factor. The proportion of females survived in the Titanic was very higher than the proportion of males.

Passenger Class: First Class passengers had the highest survival rate compared to the second and third classes.

Age: Children (0-18) generally show a higher survival rate.


Were these patterns what you expected given the historical context?
Yes, the patterns strongly align with historical context and the "women and children first" protocol which can be seen in the Sex and Age visuals

### Part 2: Linear Probability Model

1. Estimate a linear probability model predicting survival using sex, passenger class, and age.
2. Interpret your coefficients.
3. What are the limitations of using a linear probability model in this context?

```{r}
# Estimate a linear probability model predicting survival using sex, passenger class, and age.
model <- lm(Survived ~ Sex + Pclass + Age, data = titanic)
summary(model)
stargazer(model,
          type = "text",
          title = "Linear Probabilty Model (LPM) Results",
          dep.var.labels = "Survived (0 = No, 1 = Yes)",
          covariate.labels = c("Sex (Male)", "Passenger Class (2nd)", "Passenger Class (3rd)", "Age")
          )
```
Sex (male) = -0.495 
Being Male decreases the probability of survival by about 50 percentage points, compared to being Female, holding class and age constant.

Pclass2 = -0.192
Being in 2nd Class decreases the probability of survival by about 19.2 percentage points, compared to being in 1st Class, holding sex and age constant.

Pclass3 = -0.29
Being in 3rd Class decreases the probability of survival by about 29 percentage points, compared to being in 1st Class, holding sex and age constant.

Age = -0.005
Each additional one year of age decreases the probability of survival by about 0.5 percentage points, holding sex and class constant.


### Part 3: Logit Model 

1. Estimate a logit model using the same variables.


```{r}
model_1 <- glm(Survived ~ Sex + Pclass + Age, data= titanic, family = binomial(link = "logit") )
summary(model_1)

stargazer(model_1, 
          type = "text", 
          title = "Logit Model Results (Log-Odds)",
          dep.var.labels = "Survived (0=No, 1=Yes)",
          notes = "Coefficients are interpreted in log-odds.")

# Odds ratios
odds_ratios <- exp(coef(model_1))
odds_ratios_table <- data.frame(
  Variable = names(odds_ratios),
  Odds_Ratio = as.numeric(odds_ratios)
)
print("--- Odds Ratio Table ---")
print(odds_ratios_table)
```
2. Calculate and interpret the odds ratios.


```{r}
# Exponentiate the coefficients to get the Odds Ratios (e^beta)
odds_ratios <- exp(coef(model_1))

# Format and display the Odds Ratios
odds_ratios_table <- data.frame(
  Variable = names(odds_ratios),
  Odds_Ratio = as.numeric(odds_ratios)
)

print("--- Odds Ratio Table ---")
print(odds_ratios_table)
```
Sex(Male) = 0.075
Interpretation: Being Male reduces the odds of survival by a factor of approximately 0.075, compared to being female. Stated differently, the odds of a male surviving are only 7% of the odds of a female surviving.

Pclass = 0.292
Interpretation: Being in other Class reduces the odds of survival by a factor of 0.292, or a 69% (1- 0.292)x100 reduction , compared to being in 1st Class.

Age = 0.96
Interpretation: Each additional one year of age reduces the odds of survival by a factor of 0.96, or a 4% reduction
```{r}
```


```{r}
```

### Part 4: Policy Analysis 

1. Based on your models, what were the key determinants of survival on the Titanic?
2. What do your results suggest about social inequality in 1912?

Based on both models, the key determinants of survival were, in order of strength: Sex and Passenger Class.

The results implies an idea of social inequality in 1912:

The disaster protocol was supposed to be "women and children first," but our models show it was heavily influenced by "money first." The wealthy (1st Class) were located closer to the lifeboats and had better access, giving them a huge advantage. 

The steep drop in survival odds for 3rd Class passengers proves that where you stood in the social hierarchy directly impacted your opportunity to live or die. The tragedy was not just a disaster of nature, but a disaster of structural inequality.
